Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)

### 2022-04-18 doc/manuscript/ESSDmanuscript.Rmd
work on recommendations section

#Figures
fig_variable_mapping=1
fig_map=2
fig_dominant_veg=3

#Cited sections (listed here because these can change)
section_C_pools = 2.1 
section_land_use_categories = 2.2
section_new_variables = 3.2
section_C_variable_mapping = "4.1.1"
section_land_use_mapping = "4.1.2"

@@

To improve the data accessible for C accounting, the IPCC created the Emission Factor Database (EFDB; [https://www.ipcc-nggip.iges.or.jp/EFDB/main.php](https://www.ipcc-nggip.iges.or.jp/EFDB/main.php)), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.
The EFDB can be used both for efforts to tally a nation's intended or accomplished greenhouse gas reductions, or as a basis of comparison for external parties to evaluate these inventories. 
The EFDB encourages researchers to submit estimates of emission factors or other related parameters from peer-reviewed journal papers or other accepted sources for inclusion in the database. 
In the case of forests, emission factors include carbon stocks, increments ("stock changes"), and fluxes ("gains" and "losses") for various pools [@ipcc_2006_2006;@ipcc_2019_2019].

@@

The end goal of IPCC greenhouse gas inventories is to quantify greenhouse gas emissions to, or withdrawals from, the atmosphere on an annual basis, most commonly on a national level [@ipcc_2006_2006;@ipcc_2019_2019]. 
For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; t C yr^-1^) are calculated as the sum of changes in various pools (section `r section_C_pools`), plus any harvested wood products.
For each pool, $\Delta C$ may be calculated using the "Gain-Loss Method", which takes the difference between gains and losses<!--IPCC eq.2.4-->, or using the "Stock-Difference Method", which computes $\Delta C$ based on C stocks at two points in time<!--IPCC eq.2.5--> [@ipcc_2006_2006]. 
Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, increments, and fluxes in the IPCC-defined pools.

@@

As of `r format(Sys.time(), '%B %d, %Y')`, ForC (v4.0) contained `r n_independent_records_ForC` independent records (`r n_records_ForC` total), `r n_records_EFDB_variables` of which were for variables relevant to EFDB (Fig. `r fig_variable_mapping`).
These records were distributed across all forested continents and ecozones, with particularly high concentrations in *[ecozone(s)]* and low concentrations in *[ecozone(s)]* (Fig. `r fig_map`).
The most widely represented forest type was *[type]*, followed by *[type]* and *[type]* (Fig. `r fig_dominant_veg`).

@@

```{r fig_dominant_veg, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types. For FAO ecozones (b), codes are as listed in the caption of Figure 2."}
#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
knitr::include_graphics("figures_tables/Histogram_n_Relevant_and_Transferred_Records.png")
```

@@

As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, we had reviewed and sent `r n_records_sent_to_EFDB` records to EFDB, `r n_records_posted_EFDB` of which have been reviewed, accepted, and posted (Fig. `r fig_map`-`r fig_dominant_veg`, Table `r table_variables`).
*[DETAILS]*

@@

(*strongly flag both useful variables that the EFDB does not track and useful variables that papers fail to include that EFDB needs*)

@@

*(Paragraph highlighting important gaps in variables / regions)*

Several variables of value to IPCC, including standing dead wood, woody mortality, delta.agb, are not calculated and presented as frequently as are AGB and ANPP_woody, even though they can readily be derived from the same census data. 
We recommend that researchers calculate and report these, following the reporting guidelines specified below.
Furthermore, there is an opportunity to fill data gaps by calculating these from existing census data.
For example, the core census protocol of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfsforestgeo_2015,@davies_forestgeo_2021] collects the data required to calculate standing dead wood, woody mortality, and delta.agb, but these have not been calculated and reported for all sites for which the appropriate number of censuses are available <!--"(n=1 for standing dead wood, n=2 for woody mortality and delta.agb)" -- previous text commented out because I think it's out of date-->[but see @piponiot_distribution_2022, @refs<!--insert other refs including these variables-->].

Given widespread trends of increasing tree mortality[@refs], including through severe natural disturbance [@refs], better characterization of dead wood will be critical.

@@

There are plenty of relevant, published data that are not included in ForC. 
Systematic review of the literature could vastly improve data coverage.
*(There are some efforts underway, including a few that Susan can specify.)*

## IPCC protocol considerations

>>>>

#Figures
fig_variable_mapping=1
fig_map=2
fig_histograms=3

#Cited sections (listed here because these can change)
section_C_pools = 2.1 
section_land_use_categories = 2.2
section_new_variables = 3.2
section_C_variable_mapping = "4.1.1"
section_land_use_mapping = "4.1.2"
section_database_needs = "6.1"
section_new_data_needs = "6.2"
section_data_reporting_needs = "6.3" 
section_mismatches = 6.4

@@

To improve the data accessible for C accounting, the IPCC created the Emission Factor Database (EFDB; [https://www.ipcc-nggip.iges.or.jp/EFDB/main.php](https://www.ipcc-nggip.iges.or.jp/EFDB/main.php)), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.
The EFDB can be used both for efforts to tally a nation's intended or accomplished greenhouse gas reductions, or as a basis of comparison for external parties to evaluate these inventories. 
The EFDB encourages researchers to submit estimates of emission factors or other related parameters from peer-reviewed journal papers or other accepted sources for inclusion in the database. 
In the case of forests, emission factors include carbon stocks, net increments ("stock changes"), and fluxes ("gains" and "losses") for various pools [@ipcc_2006_2006;@ipcc_2019_2019].

@@

The end goal of IPCC greenhouse gas inventories is to quantify greenhouse gas emissions to, or withdrawals from, the atmosphere on an annual basis, most commonly on a national level [@ipcc_2006_2006;@ipcc_2019_2019]. 
For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; t C yr^-1^) are calculated as the sum of changes in various pools (section `r section_C_pools`), plus any harvested wood products.
For each pool, $\Delta C$ may be calculated using the "Gain-Loss Method", which takes the difference between gains and losses<!--IPCC eq.2.4-->, or using the "Stock-Difference Method", which computes $\Delta C$ based on C stocks at two points in time<!--IPCC eq.2.5--> [@ipcc_2006_2006]. 
Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, net annual increments, and fluxes in the IPCC-defined pools.

@@

As of `r format(Sys.time(), '%B %d, %Y')`, ForC (v4.0) contained `r n_independent_records_ForC` independent records (`r n_records_ForC` total), `r n_records_EFDB_variables` of which were for variables relevant to EFDB (Fig. `r fig_variable_mapping`).
These records were distributed across all forested continents and ecozones, with particularly high concentrations in *[ecozone(s)]* and low concentrations in *[ecozone(s)]* (Fig. `r fig_map`).
The most widely represented forest type was *[type]*, followed by *[type]* and *[type]* (Fig. `r fig_histograms`).

@@

```{r fig_histograms, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types. For FAO ecozones (b), codes are as listed in the caption of Figure 2."}
#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
knitr::include_graphics("figures_tables/Histogram_n_Relevant_and_Transferred_Records.png")
```

@@

As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, we had reviewed and sent `r n_records_sent_to_EFDB` records to EFDB, `r n_records_posted_EFDB` of which have been reviewed, accepted, and posted (Figs. `r fig_map`-`r fig_histograms`, Table `r table_variables`).
*[DETAILS]*

@@

Based on our experience contributing forest C data to EFDB via ForC, we make several recommendations as to how scientists can improve forest C records in EFDB through database work (section `r section_database_needs`), new data collection and analysis (section `r section_new_data_needs`), and reporting (section `r section_data_reporting_needs`).
We also highlight notable mismatches between IPCC accounting methods and forest C mensuration (section `r section_mismatches`).

There is vast potential to expand forest C data in EFDB by completing the process of reviewing and sending data that are already in ForC (Figs. `r fig_map`-`r fig_histograms`).
So far, only ~`r round(n_records_sent_to_EFDB/n_records_EFDB_variables,2)*100`% of the EFDB-relevant data in ForC have been sent to EFDB.

Moreover, there are many published EFDB-relevant forest C data that are not included in ForC, with more being published on a nearly daily basis. 
Coverage of particular variables or regions could be vastly improved through systematic review of the literature.
*(There are some efforts underway, including a few that Susan can specify.)*
Such reviews are necessary to even develop a rigorous assessment of forest C data that are available, versus those that require additional data collection and analysis.

@@

New data collection and analysis is needed to fill notable knowledge gaps.
While aboveground biomass stocks in particular have received -- and continue to receive -- significant research attention, other pools and variables remain poorly quantified [Table `r table_variables`, @anderson-teixeira_carbon_2021].
Furthermore, data distribution is uneven across forest types and geographical regions (Figs. `r fig_map`-`r fig_histograms`).
For instance, C cycling of tropical forests -- particularly in Africa -- remains relatively poorly characterized, in large part due to substantial barriers to data collection and distribution [@delima_making_2022]
*(add some more here?)*

Several variables of value for IPCC C accounting have not been calculated and presented as would be possible given the same forest census data and minimal extra research effort.
For example, aboveground woody mortality (*woody.mortality_ag*) and aboveground biomass increment (*delta.agb*) can be calculated from the same census data as aboveground woody productivity (*ANPP_woody*), yet the latter has received far more research attention, and correspondingly has far more records in ForC [Table `r table_variables`, @anderson-teixeira_carbon_2021; but see @piponiot_distribution_2022]. 
Similarly, live coarse root biomass, total biomass, and changes in both of these pools could in theory be easily be estimated in parallel with aboveground biomass, with the greatest barrier being availability of reliable allometries, as have been developed for aboveground biomass [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022].
However, while equations for estimating root (and thereby total) biomass require improvement, they do exist *for many forest types*[@refs], and IPCC provides default recommendations of below-ground to above-ground ratios for estimation of root biomass [@ipcc_2019_2019].
In addition, standing dead trees are captured in most forest censuses and could be used to estimate standing dead wood, although additional data on breakage would be needed for accurate accounting. 
We recommend that, when possible, researchers calculate these, following the reporting guidelines specified in section `r section_data_reporting_needs`.

*Other EFDB-relevant variables require more effort but are warrented given their importance for forest C accounting.*
Given widespread trends of increasing tree mortality[@refs], including through severe natural disturbance [@refs], better characterization of dead wood will be critical.
...

@@

## Mismatches between IPCC accounting methods and forest C mensuration

IPCC accounting methods cannot leverage eddy-covariance measurements, which are widely seen as the best available method for quantifying ecosystem-atmosphere gas exchange.

### 2022-04-18 doc/manuscript/ESSDmanuscript.tex
work on recommendations section

To improve the data accessible for C accounting, the IPCC created the
Emission Factor Database (EFDB;
\url{https://www.ipcc-nggip.iges.or.jp/EFDB/main.php}), which is
intended as a recognized library of emission factors and other
parameters that can be used for estimating greenhouse gas emissions and
removals. The EFDB can be used both for efforts to tally a nation's
intended or accomplished greenhouse gas reductions, or as a basis of
comparison for external parties to evaluate these inventories. The EFDB
encourages researchers to submit estimates of emission factors or other
related parameters from peer-reviewed journal papers or other accepted
sources for inclusion in the database. In the case of forests, emission
factors include carbon stocks, increments (``stock changes''), and
fluxes (``gains'' and ``losses'') for various pools
\citep{ipcc_2006_2006, ipcc_2019_2019}.

@@

The end goal of IPCC greenhouse gas inventories is to quantify
greenhouse gas emissions to, or withdrawals from, the atmosphere on an
annual basis, most commonly on a national level
\citep{ipcc_2006_2006, ipcc_2019_2019}. For each stratum of subdivision
within a land-use category, annual stock changes (\(\Delta C\); t C
yr\textsuperscript{-1}) are calculated as the sum of changes in various
pools (section 2.1), plus any harvested wood products. For each pool,
\(\Delta C\) may be calculated using the ``Gain-Loss Method'', which
takes the difference between gains and losses, or using the
``Stock-Difference Method'', which computes \(\Delta C\) based on C
stocks at two points in time \citep{ipcc_2006_2006}. Thus, C cycle
variables relevant to the IPCC methodology and to EFDB include C stocks,
increments, and fluxes in the IPCC-defined pools.

@@

\begin{figure}
\includegraphics[width=15cm]{figures_tables/Histogram_n_Relevant_and_Transferred_Records} \caption{\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types. For FAO ecozones (b), codes are as listed in the caption of Figure 2.}\label{fig:fig_dominant_veg}
\end{figure}

@@

As of April 18, 2022, we had reviewed and sent NA records to EFDB, 73 of
which have been reviewed, accepted, and posted (Fig. 2-3, Table 2).
\emph{{[}DETAILS{]}}

@@

(\emph{strongly flag both useful variables that the EFDB does not track
and useful variables that papers fail to include that EFDB needs})

@@

\emph{(Paragraph highlighting important gaps in variables / regions)}

Several variables of value to IPCC, including standing dead wood, woody
mortality, delta.agb, are not calculated and presented as frequently as
are AGB and ANPP\_woody, even though they can readily be derived from
the same census data. We recommend that researchers calculate and report
these, following the reporting guidelines specified below. Furthermore,
there is an opportunity to fill data gaps by calculating these from
existing census data. For example, the core census protocol of the
Forest Global Earth Observatory {[}ForestGEO;
\citet{anderson-teixeira_ctfsforestgeo_2015},\citet{davies_forestgeo_2021}{]}
collects the data required to calculate standing dead wood, woody
mortality, and delta.agb, but these have not been calculated and
reported for all sites for which the appropriate number of censuses are
available \citep[but see][\citet{refs}]{piponiot_distribution_2022}.

Given widespread trends of increasing tree mortality\citep{refs},
including through severe natural disturbance \citep{refs}, better
characterization of dead wood will be critical.

@@

There are plenty of relevant, published data that are not included in
ForC. Systematic review of the literature could vastly improve data
coverage. \emph{(There are some efforts underway, including a few that
Susan can specify.)}

\subsection{IPCC protocol considerations}

>>>>

To improve the data accessible for C accounting, the IPCC created the
Emission Factor Database (EFDB;
\url{https://www.ipcc-nggip.iges.or.jp/EFDB/main.php}), which is
intended as a recognized library of emission factors and other
parameters that can be used for estimating greenhouse gas emissions and
removals. The EFDB can be used both for efforts to tally a nation's
intended or accomplished greenhouse gas reductions, or as a basis of
comparison for external parties to evaluate these inventories. The EFDB
encourages researchers to submit estimates of emission factors or other
related parameters from peer-reviewed journal papers or other accepted
sources for inclusion in the database. In the case of forests, emission
factors include carbon stocks, net increments (``stock changes''), and
fluxes (``gains'' and ``losses'') for various pools
\citep{ipcc_2006_2006, ipcc_2019_2019}.

@@

The end goal of IPCC greenhouse gas inventories is to quantify
greenhouse gas emissions to, or withdrawals from, the atmosphere on an
annual basis, most commonly on a national level
\citep{ipcc_2006_2006, ipcc_2019_2019}. For each stratum of subdivision
within a land-use category, annual stock changes (\(\Delta C\); t C
yr\textsuperscript{-1}) are calculated as the sum of changes in various
pools (section 2.1), plus any harvested wood products. For each pool,
\(\Delta C\) may be calculated using the ``Gain-Loss Method'', which
takes the difference between gains and losses, or using the
``Stock-Difference Method'', which computes \(\Delta C\) based on C
stocks at two points in time \citep{ipcc_2006_2006}. Thus, C cycle
variables relevant to the IPCC methodology and to EFDB include C stocks,
net annual increments, and fluxes in the IPCC-defined pools.

@@

\begin{figure}
\includegraphics[width=15cm]{figures_tables/Histogram_n_Relevant_and_Transferred_Records} \caption{\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types. For FAO ecozones (b), codes are as listed in the caption of Figure 2.}\label{fig:fig_histograms}
\end{figure}

@@

As of April 18, 2022, we had reviewed and sent NA records to EFDB, 73 of
which have been reviewed, accepted, and posted (Figs. 2-3, Table 2).
\emph{{[}DETAILS{]}}

@@

Based on our experience contributing forest C data to EFDB via ForC, we
make several recommendations as to how scientists can improve forest C
records in EFDB through database work (section 6.1), new data collection
and analysis (section 6.2), and reporting (section 6.3). We also
highlight notable mismatches between IPCC accounting methods and forest
C mensuration (section 6.4).

There is vast potential to expand forest C data in EFDB by completing
the process of reviewing and sending data that are already in ForC
(Figs. 2-3). So far, only \textasciitilde NA\% of the EFDB-relevant data
in ForC have been sent to EFDB.

Moreover, there are many published EFDB-relevant forest C data that are
not included in ForC, with more being published on a nearly daily basis.
Coverage of particular variables or regions could be vastly improved
through systematic review of the literature. \emph{(There are some
efforts underway, including a few that Susan can specify.)} Such reviews
are necessary to even develop a rigorous assessment of forest C data
that are available, versus those that require additional data collection
and analysis.

@@

New data collection and analysis is needed to fill notable knowledge
gaps. While aboveground biomass stocks in particular have received --
and continue to receive -- significant research attention, other pools
and variables remain poorly quantified \citep[Table
2,][]{anderson-teixeira_carbon_2021}. Furthermore, data distribution is
uneven across forest types and geographical regions (Figs. 2-3). For
instance, C cycling of tropical forests -- particularly in Africa --
remains relatively poorly characterized, in large part due to
substantial barriers to data collection and distribution
\citep{delima_making_2022} \emph{(add some more here?)}

Several variables of value for IPCC C accounting have not been
calculated and presented as would be possible given the same forest
census data and minimal extra research effort. For example, aboveground
woody mortality (\emph{woody.mortality\_ag}) and aboveground biomass
increment (\emph{delta.agb}) can be calculated from the same census data
as aboveground woody productivity (\emph{ANPP\_woody}), yet the latter
has received far more research attention, and correspondingly has far
more records in ForC \citetext{\citealp[Table
2,][]{anderson-teixeira_carbon_2021}; \citealp[but
see][]{piponiot_distribution_2022}}. Similarly, live coarse root
biomass, total biomass, and changes in both of these pools could in
theory be easily be estimated in parallel with aboveground biomass, with
the greatest barrier being availability of reliable allometries, as have
been developed for aboveground biomass
\citep{chave_improved_2014, rejou-mechain_biomass_2017, gonzalez-akre_allodb_2022}.
However, while equations for estimating root (and thereby total) biomass
require improvement, they do exist \emph{for many forest
types}\citep{refs}, and IPCC provides default recommendations of
below-ground to above-ground ratios for estimation of root biomass
\citep{ipcc_2019_2019}. In addition, standing dead trees are captured in
most forest censuses and could be used to estimate standing dead wood,
although additional data on breakage would be needed for accurate
accounting. We recommend that, when possible, researchers calculate
these, following the reporting guidelines specified in section 6.3.

\emph{Other EFDB-relevant variables require more effort but are
warrented given their importance for forest C accounting.} Given
widespread trends of increasing tree mortality\citep{refs}, including
through severe natural disturbance \citep{refs}, better characterization
of dead wood will be critical. \ldots{}

@@

\subsection{Mismatches between IPCC accounting methods and forest C
mensuration}

IPCC accounting methods cannot leverage eddy-covariance measurements,
which are widely seen as the best available method for quantifying
ecosystem-atmosphere gas exchange.
